# Reproducible Research: Peer Assessment 1


## 1. Loading and preprocessing the data
Set the working directory:
```{r setWorkingDir}
setwd("~/Documents/Coursera/dataScienceSpecialisation/RepData_PeerAssessment1")
```

Read in the data:
```{r readData}
data <- read.csv("activity.csv")
```

Do some transformations to get the data ready for analysis:
```{r transform}
require(data.table)
```

```{r groupBy}
## Get Steps per day
## group data by 'date' and find the mean of 'steps'
data <- data.table(data)
dt <- data[,list(meanSteps=mean(steps, na.rm=TRUE), std=sd(steps, na.rm=TRUE), totalPerDay=sum(steps, na.rm=TRUE)), by=date]
timeseries <- data[,list(meanSteps=mean(steps, na.rm=TRUE), std=sd(steps, na.rm=TRUE), totalPerDay=sum(steps, na.rm=TRUE)), by=interval]
```

## 2. What is mean total number of steps taken per day?
### 2.1 Histogram of total number of steps taken each day:
```{r makeHist1}
require(ggplot2)

# Histogram: Total Steps Per Day
p <- ggplot(dt, aes(x=totalPerDay)) 
p <- p + geom_histogram(aes(fill=..count..))
p <- p+ scale_fill_gradient("Count", low = "red", high = "green")
p










################################
# Total Steps Per Day
p <- ggplot(dt, aes(x=date, y=totalPerDay, fill=totalPerDay)) 
p <- p + geom_bar(stat="identity", colour="darkblue")
p <- p + ylab("Total Steps per Day")
p <- p + xlab("Date")
p <- p + ggtitle("fig 2.1: Histogram - Total Number of Steps Per Day")
p <- p + theme(legend.position="none")
p

# Mean Steps Per Day
p <- ggplot(dt, aes(x=date, y=meanSteps))
p <- p + geom_bar(stat="identity", colour="darkred", fill="pink")
p <- p + ylab("Mean Steps per Day")
p <- p + xlab("Date")
p <- p + ggtitle("fig 2.2: Mean Number of Steps Per Day Over All Daily Intervals")
p <- p + theme(legend.position="none")
p

myMean <- mean(data$steps, na.rm = TRUE)
myMedianAcrossAllDays <- median(data$steps, na.rm = TRUE)
myMedianDailyAve <- median(dt$meanSteps, na.rm = TRUE)
```
  
### 2.2 The mean and median total number of steps taken per day are:  
#### Mean:       `r myMean`     
#### Median Across All Days:     `r myMedianAcrossAllDays` 
#### Median of Daily Averages: `r myMedianDailyAve`
  
  
## 3. What is the average daily activity pattern?
### 3.1 Time Series Plot
```{r makeTimeSeries1}
m <- ggplot(timeseries, aes(interval, meanSteps)) 
m <- m + geom_line(colour="darkblue")
m <- m + theme_bw()
m <- m + ggtitle("fig 3.1: Timeseries - Daily Activity Pattern")
m <- m + xlab("Interval") + ylab("Average Number of Steps") 
m
```

### 3.2 Which 5min Interval, on average across all days, contains maximum number of steps?
```{r findMax}
intervalWithMaxSteps <- timeseries[which(meanSteps == max(timeseries$meanSteps)), ]$interval
```
Therefore interval `r intervalWithMaxSteps` has the maximum average number of steps per day.


## 4. Imputing missing values
### 4.1 Calculate the total number of missing values in dataset.
```{r missingVals} 

numMissingVals  <- nrow(data[ complete.cases(data)==FALSE ])
numMissingVals

```
#### Total number of missing = `r numMissingVals`
  
  
### 4.2 Fill in the missing values using the min for the corresponding 5min interval
```{r fillMissingVals}
## Strategy:
## Use the mean for that 5min interval to replace the mising values

## Build a custom coalesce function
coalesce2<-function(...){
  Reduce(function(x,y) {
    i<-which(is.na(x))
    x[i]<-y[i]
    x},
  list(...))
}

## Join the original Data with the Time Series data by the common column "interval"
merged <- merge(data, timeseries, by = "interval")

## Where steps is NA in the original dataset, replace it with the mean number of steps for that interval
merged$steps <- with(merged, coalesce2(steps, meanSteps))
```


### 4.3 New dataset equal to the original dataset but with the missing data filled in
```{r newDataset}
data2 <- subset(merged, select=cbind("steps","date","interval"))
```

See head of new dataset below. _Notice there are no NAs here._
```{r printHeadNewData, echo=FALSE}
head(data2)      
```
  

### 4.4 Histogram of Total number of steps taken each day
```{r makeHist2}

dt2 <- data2[,list(meanSteps=mean(steps, na.rm=TRUE), std=sd(steps, na.rm=TRUE), totalPerDay=sum(steps, na.rm=TRUE)), by=date]

# Histogram: Total Steps Per Day
p <- ggplot(dt2, aes(x=totalPerDay)) 
p <- p + geom_histogram(aes(fill=..count..))
p <- p+ scale_fill_gradient("Count", low = "red", high = "green")
p

##############################################################
# Total Steps Per Day
p <- ggplot(dt2, aes(x=date, y=totalPerDay, fill=totalPerDay))
p <- p + geom_bar(stat="identity", colour="darkblue")
p <- p + ylab("Total Steps per Day")
p <- p + xlab("Date")
p <- p + ggtitle("Fig 4.4: Histogram - Total Number of Steps Per Day")
p <- p + theme(legend.position="none")
p


myMean2 <- mean(data2$steps, na.rm = TRUE)
myMedianAcrossAllDays2 <- median(data2$steps, na.rm = TRUE)
myMedianDailyAve2 <- median(dt2$meanSteps, na.rm = TRUE)
```

```{r, echo=FALSE}
delta <- abs(myMedianDailyAve2 - myMedianDailyAve)
```

### 4.5 The mean and median total number of steps taken per day are:  
#### Mean:       `r myMean2`     
#### Median Across All Days:     `r myMedianAcrossAllDays2` 
#### Median of Daily Averages: `r myMedianDailyAve2`

### 4.6 Do these values differ from the estimates from the first part of the assignment? 
Note the **Mean** and **Median Across All Days** remain unchanged after replacing the missing values with the mean of the corresponding day. However, the **Median of Daily Averages** differs by `r delta`. 
  
### 4.7 Effects of Imputing Missing Values on estimates of total daily number of steps
The total number of steps taken per day have increased on eight days as a result of replacing the missing values with the means. These eight days previously had NAs and now have values. This can be see in the figure below with the vertical bars representing the days where the total number of daily steps differed from the original dataset. 

#### Histogram - Difference in Total Number of Steps Per Day Resulting from Imputing Missing values
```{r }
diff <- dt2
diff$diff.totalPerDay <- dt2$totalPerDay - dt$totalPerDay

# Total Steps Per Day
p <- ggplot(diff, aes(x=date, y=diff.totalPerDay, fill=diff.totalPerDay)) 
p <- p + geom_bar(stat="identity", colour="darkblue")
p <- p + theme(legend.position="none")
p <- p + xlab("Date") + ylab("Total Difference in Steps") 
p <- p + ggtitle("fig 4.7: Difference in Total Number of Steps Per Day")
p
```

## 5. Are there differences in activity patterns between weekdays and weekends?
- Use weekdays() function
- Use dataset with filled in values

### 5.1 Create new factor in dataset with 2 levels "Weekday", "Weekend"
```{r makeWeekdays}
data2$date <- as.Date(data2$date)
data2$weekdayName  <- with(data2, weekdays(date))
data2$weekdayFlag <- factor(with(data2, ifelse(weekdays(date) %in% c("Monday","Tuesday", "Wednesday","Thursday", "Friday")  ,"Weekday" ,"Weekend" )) )

timeseries2 <- data2[,list(meanSteps=mean(steps, na.rm=TRUE), std=sd(steps, na.rm=TRUE), totalPerDay=sum(steps, na.rm=TRUE)), by=c("weekdayFlag","interval")]

```


### 5.2 Make time series panel plot (type "l") x-axis = 5min interval, y-axis = avg steps taken
```{r makeTimeSeries2}

m <- ggplot(timeseries2, aes(interval, meanSteps)) 
m <- m + geom_line(colour="darkblue") 
m <- m + facet_wrap(~weekdayFlag, nrow=2, ncol=1)
m <- m + xlab("Interval")
m <- m + ylab("Number of Steps")
m <- m + ggtitle("fig 5.2: 5min intervals vs average number of steps per interval")
m <- m + theme_bw()
m <- m + theme(strip.text.x = element_text(size=14, face="bold"),
          strip.background = element_rect(colour="black", fill="skyblue"))
m
```

```{r makeTimeSeries3, echo=FALSE}

m <- ggplot(timeseries2, aes(interval, meanSteps)) 
m <- m + geom_line(colour="darkblue") 
m <- m + geom_smooth(alpha=.2, size=1, colour="red")
m <- m + facet_wrap(~weekdayFlag, nrow=2, ncol=1)
m <- m + xlab("Interval")
m <- m + ylab("Number of Steps")
m <- m + ggtitle("fig 5.3: 5min intervals vs average number of steps per interval")
m <- m + theme_bw()
m <- m + theme(strip.text.x = element_text(size=14, face="bold"),
          strip.background = element_rect(colour="black", fill="skyblue"))
m
```

```{r makeDeltaPlot, echo=FALSE}
# ## Split into weekdays data and weekends Data
# weekdaysData <-  timeseries2[which(weekdayFlag == "Weekday"),]
# weekendData <- timeseries2[which(weekdayFlag == "Weekend"),]
# 
# ## Merge datasets
# weekdayVSWeekend <- merge(weekdaysData,weekendData, by="interval")
# weekdayVSWeekend$meanStepsDelta <- with(weekdayVSWeekend, meanSteps.x - meanSteps.y)
# 
# m <- ggplot(weekdayVSWeekend, aes(interval, meanStepsDelta)) 
# m <- m + geom_point() 
# m <- m + geom_smooth(alpha=.2, size=1)
# m <- m + xlab("Interval")
# m <- m + ylab("Number of Steps")
# m <- m + ggtitle("fig 5.4: Difference in Activity Between Weekdays and Weekends")
# m
# **++Note Weekdays are Positive Values and Weekends are Negative Values** 
```
  
 
  
#### Conculsion
- There are there differences in activity patterns between weekdays and weekends. Activity tends to start later and more gradually on weekends than on weekdays according to plot 5.2. For weekends it starts from interval 500 onwards very gradually. Whereas for weekdays it begins from interval 500 onwards very abruptly.

- For weekdays activity tend to be more periodic with a spike every few hours, whereas weekends do not exhibit this periodic nature. 

- Weekend activity tends to cease later in the day with a large spike just after interval 2000. Whereas weekday activity more abrubtly ends just after interval 1550




















